Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-16 docs/proposals/NSF/ProjDesc/ProjDesc.tex
worked on later sections and added a few comments throughout

The first problem discussed here is the erosion of a porous medium, as has been preliminarily investigated in~\cite{Quaife2018}. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, erosion results from surface shear stress
\begin{align}
\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss \, .
\end{align}
where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations \eqref{StokesEq}. 
The motion of the interface is proportional to the absolute value of shear
\begin{align}
\Vn = \CE \, \abs{\tau},
\end{align}
In the latter case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
\label{Ceq}
& \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C \, ,
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \left. \Diff \pd{C}{\ny} \right. %\right\vert_{\bdry} \, ,
\end{equation}
Both processes have been investigated by the PIs and collaborators in
the context of single bodies in high Reynolds-number
flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In
the low-Reynolds regime of porous-media flow, numerical methods have
been developed by the PIs to simulate shear-induced
erosion~\cite{Quaife2018}. These simulations combine highly accurate
boundary integral methods with stable interface evolution techniques.
The boundary integral solver is accelerated by the fast multipole
method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$
complexity. Meanwhile, the so-called {\thL} method is used for interface
evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.
%A near-term goal is to adapt these methods to handle the case of chemically-induced dissolution.

@@

 Figure \ref{fig1} shows preliminary results of multiple bodies undergoing mechanical erosion in Stokes flow. The figure shows 80 bodies of initially random size and position, undergoing erosion. This figure extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015}, which enables closer contact between bodies and thus more bodies.
The color shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity. Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. The channelization process is made more evident by visualization of the velocity magnitude, as in Fig.~\ref{fig2}.

@@

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows new, preliminary results on how media properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies erode. 
%Time is normalized by the value $t_f$ at which all bodies vanish and hence the area fraction is zero. 
Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation \eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially. The much more mild decrease in area fraction is not sufficient to account for this exponential rate, and hence the reshaping process, in particular the formation of channels, plays a pivotal role.

@@

Though conceptually straightforward, implementing contact forces in multibody fluid-structure simulations is far from simple in practice, with several competing possible strategies that are a topic of current research.  One approach is to use introduce a potential with short-range repulsion, but such methods often introduce stiffness. Alternatively, a space-time interference volume can guarantee a minimum separation between grains.  In a fluid context, this method was first introduced by Lu et al.~\cite{lu-rah-zor2017}, and extended to allow for much smaller separation distances of rigid bodies by the PI's previous graduate student~\cite{bys-sha-qua2019}. One Ph.D.~student, who will be jointly supervised by both PIs, will simulate sedimentation of eroding bodies by developing algorithms that resolve quadrature error and contact.

@@

\begin{wrapfigure}[]{r}{0.3\textwidth}
  \includegraphics[width=0.3\textwidth]{figs/100b_t100tracer}
  \caption{\label{fig:100tracers} A visualization of the streamlines
  with 200 tracers passing through an eroded geometry with a porosity of
  0.63. The tracers are initialized at the left end of the channel and
  the flow is driven by a pressure difference from left to right.}
\end{wrapfigure}

@@

In an eroded geometry $\Omega$, the dimensionless
concentration $c$ is governed by
\begin{align}
  \pd{c}{t} + \uu \cdot \nabla c = \frac{1}{\Pe} \Delta c, 
  \label{eqn:advectionDiffusion}
\end{align}
where $\Pe$ is the dimensionless Peclet number. Here we assume that the
concentration $c$ is affected by the eroded geometry, but the geometry
is not altered further by the concentration. In the infinite Peclet
number limit, the concentration is governed by pure advection and the
characteristics of the concentration field are identical to the
streamlines (Figure~\ref{fig:100tracers}).

The $j^{th}$ trajectory, or characteristic, is governed by $\dot{\ss}_j(t) = \uu(\ss_j(t))$. Once these trajectories are computed, using for example a high-order Runge-Kutta method, a statistical analysis can be used to describe the transport. Two of the most important measurements are the tortuosity and the anomalous dispersion, both which are defined in terms of the length of the trajectories \begin{align}
  \lambda_j(t) = \int_{0}^{t} \|\ss'(t')\| dt'.
\end{align}
Upon computing $N_p$ trajectories, the tortuosity is the dimensionless number
\begin{align}
  T = \frac{1}{d} \left(\int_{S} u_1(x_0,y_0)\lambda(y_0) dy_0\, \right)
  \Bigg/ \left(\int_S u_1(x_0,y_0)\, dy_0 \right),
\end{align}
where $S$ is the vertical cross-section at $x=x_0$, $d$ is its length,
$u_1$ is the horizontal velocity, and $\lambda(y_0)$ is the length of
the trajectory starting at $(x_0,y_0)$ and ending at a vertical
cross-section at the other end of the channel. Relationships between the
tortuosity and porosity of a geometry have been developed by other
groups~\cite{kop-kat-tim1996, dud-koz-mat2011, mat-kha-koz2008}, and
preliminary results indicate that such models might be appropriate for
geometries whose porosity is increasing under the process of erosion
(Figure~\ref{fig:100tortuosity}(a)). However, preliminary results also
indicate that erosion can increase the amount of tortuosity as is
evident towards the at the towards the end of the erosion process in
Figure~\ref{fig:100tortuosity}(a). Moreover, we have observed a
significant increase in the tortuosity at small porosities
(Figure~\ref{fig:100tortuosity}(b)).

\begin{figure}[htp]
  \begin{center}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian100}
  \caption{100 eroding bodies}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian20}
  \caption{20 eroding bodies}
  \end{subfigure}
  \end{center}
  \caption{\label{fig:100tortuosity} The tortuosity of two different eroding geometries. The tortuosity can be computed using two different formulas (red vs.~blue marks). In the left plot, the red square is the tortuosity of the geometry in Figure~\ref{fig:100tracers}. In both cases, there is qualitative agreement between our numerical simulations and a power-law model (black dashed line)~\cite{mat-kha-koz2008}.}
\end{figure}

While the tortuosity characterizes the amount of winding of streamlines, the dispersion characterizes the spreading of the trajectories.  Dispersion can be described using the first- and second-ensemble moments of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \lambda_j(t), \quad \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2
\end{align}
Then, the standard deviation $\sigma_\lambda$ characterizes the dispersion, and long-time behavior typically results in a power-law scaling $\sigma_\lambda \sim t^{\alpha}$. The power $\alpha$, which quantifies the long-time spreading, is $1$ in a ballistic regime which occurs at early times when the trajectories have not explored much of the porous geometry. However, once the trajectories have traversed a few grains, dispersion in porous media is often super-dispersive with $\alpha \in (0.5,1)$~\cite{dea-leb-den-tar-bol-dav2013}. In Figure~\ref{fig:100dispersion}, we investigate spreading at several different porosities of a porous geometry that is initialized with 100 grains.
\begin{wrapfigure}[22]{r}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} The particle spreading,
  characterized with the standard deviation $\sigma_\lambda(t)$ at 6
  different porosities. At early times, the line of slope 1 (black
  curve) indicates a ballistic motion. At later times, the
  dispersion becomes super-dispersive at a rate indicated by the slope
  of the black dashed-dotted line.}
\end{wrapfigure}

The rate of dispersion is known to be directly related to the pore size
distribution~\cite{dea-qua-bir-jua2018}. Therefore, there is a need to
understand how erosion affects the pore size distribution. In
preliminary work~\cite{chi-moo-qua2019}, we define pore sizes between grains that share an edge of a Delanuay triangulation with vertices at the grain centers (Figure~\ref{fig:Eroding100gap_mean_var}(a)). In Figures~\ref{fig:Eroding100gap_mean_var}(a) and~\ref{fig:Eroding100gap_mean_var}(b), we plot the mean and variance of the pore sizes as a function of porosity. To better understand how erosion affects the anomalous dispersion, we propose to investigate how the pores grow with erosion. We will start with the notion of a {\em porelet} which is an individual Hagen-Poiseuille flow in each of the individual pores~\cite{dea-qua-bir-jua2018}. By computing the strength of the individual Hagen-Poiseuille flows, the shear rate can be computed, and thus the rate of erosion can be computed.

We have performed a carefully chosen simulation to better understand the
effect of erosion on the pore size. We start with a single grain inside
the channel that is sufficiently large that the spacing between the
initial grain and the channel walls, which is 2 units wide), is
$10^{-3}$. Since we use a Barycentric quadrature rule, we are able to
stably simulate erosion with only 512 points on the eroding body which
corresponds to an arclength spacing 12 times larger than the original
separation distance---to achieve a comparable accuracy with the spectral
trapezoid rule, 60 times as many discretization points would be
required. In Figure~\ref{fig:porelets}(a), different shades of grey
correspond to the configuration of the eroding grain at five equally
spaced time steps. At three of the time steps, we plot the velocity
between the grain and the solid wall (black line). At all three
configurations, the flow is nearly parabolic indicating that  porelet
model~\cite{dea-qua-bir-jua2018} is appropriate. After performing a
thorough analysis of the pore size growth in this simple configuration,
we will develop a more complete model for networks of pores that are
prevalent in porous media.

@@

These simulation have not yet begun to incorporate the slow-acting effects of mechanical and chemical erosion, and the associated gradual changes in medium properties. A grand goal of this proposal would be to merge the two lines of inquiry to develop a sound theoretical framework for porous-medium dynamics robust enough to capture both short-term events, such as sinkhole collapse, and long-term changes in the medium. The greatest challenge faced is the disparate temporal and spatial scales that must be considered. Our proposed strategy is to use the high-fidelity erosion simulations to parametrize the effect of erosion on medium properties.

@@

\todo[inline]{Bryan is here}
\subsection{Undergraduate research}
Regarding undergraduate research, the ideas in this proposal represent a rich and fascinating topic for undergraduate students, while also being approachable these students. In particular, we aim to recruit at least one undergraduate student to perform laboratory experiments at the Geophysical Fluid Dynamics Institute (GFDI). As mentioned above, GFDI offers state-of-the-art laboratory facilities, with an already-functioning sinkhole experiment. The student will perform controlled experiments using natural and artificial porous media (e.g.~sand and micro-glass beads respectively) to precisely characterize the parameter regime leading to gravitational collapse. Parallel experiments will be conducted with erodible (bentonite clay) or dissolvable  (sucrose solids) fragments materials, which will represent the porous medium. These experiments will be used to guide, calibrate, and validate the computational advances. These undergraduate students will work closely with the graduate students to ensure that theory and experiments advance in close cooperation, with constant feedback between the two.

PI Moore already has experience successfully training undergraduate students in combined laboratory and theoretical work. A recent example is undergraduate student Tyler Bolles who completed an honors thesis with PI Moore at FSU on laboratory and theory of water waves. This work resulted in one {\em peer-reviewed} publication already~\cite{Bolles2019}, with a second to be submitted soon.

@@

\subsection{Outreach for K--12 and the general public}

\todo[inline]{FSU's YSP}
The PIs have already seized the opportunity to engage K-12 students in outreach activities. In particular, Moore's research was the focus of an educational video created by {\bf CPALMS}, which is the State of Florida's official source for standards information and course descriptions for K-12 education. The video, which features an interview and animations of research, is used to reinforce concepts from mathematics courses (grade levels 7-12), and to encourage students to consider a career in the STEM fields~\cite{CPALMS}. Additionally, these videos are frequently used for the continuing education of K-12 {\em teachers}, providing them a broader perspective of mathematics that could be integrated into their teaching.

@@

>>>>

The first problem discussed here is the erosion of a porous medium, as has been preliminarily investigated in~\cite{Quaife2018}. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, erosion results from surface shear stress
\begin{align}
\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss \, .
\end{align}
where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{StokesEq}. 
The motion of the interface is proportional to the absolute value of shear
\begin{align}
\Vn = \CE \, \abs{\tau},
\end{align}
In the latter case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
\label{Ceq}
& \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C \, ,
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \left. \Diff \pd{C}{\ny} \right. %\right\vert_{\bdry} \, ,
\end{equation}
Both processes have been investigated by the PIs and collaborators in
the context of single bodies in high Reynolds-number
flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In
the low-Reynolds regime of porous-media flow, numerical methods have
been developed by the PIs to simulate shear-induced
erosion~\cite{Quaife2018}. These simulations combine highly accurate
boundary integral methods with stable interface evolution techniques.
The boundary integral solver is accelerated by the fast multipole
method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $O(N)$
complexity. Meanwhile, the so-called {\thL} method is used for interface
evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.
%A near-term goal is to adapt these methods to handle the case of chemically-induced dissolution.

@@

Fig.~\ref{fig1} shows preliminary results of multiple bodies undergoing mechanical erosion in Stokes flow. The figure shows 80 bodies of initially random size and position, undergoing erosion. This figure extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015}, which enables closer contact between bodies and thus more bodies.
The color shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity. Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. The channelization process is made more evident by visualization of the velocity magnitude, as in Fig.~\ref{fig2}.

@@

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows new, preliminary results on how media properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies erode. 
%Time is normalized by the value $t_f$ at which all bodies vanish and hence the area fraction is zero. 
Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially. The much more mild decrease in area fraction is not sufficient to account for this exponential rate, and hence the reshaping process, in particular the formation of channels, plays a pivotal role.

@@

Though conceptually straightforward, implementing contact forces in multibody fluid-structure simulations is far from simple in practice, with several competing possible strategies that are a topic of current research.  One approach is to use introduce a potential with short-range repulsion, but such methods often introduce stiffness. Alternatively, a space-time interference volume can guarantee a minimum separation between grains.  In a fluid context, this method was first introduced by Lu et al.~\cite{lu-rah-zor2017}, and extended to allow for much smaller separation distances of rigid bodies by the PI's previous graduate student~\cite{bys-sha-qua2019}. One PhD student, who will be jointly supervised by both PIs, will simulate sedimentation of eroding bodies by developing algorithms that resolve quadrature error and contact.

@@

\begin{wrapfigure}[]{r}{0.5\textwidth}
  \includegraphics[width=0.5\textwidth]{figs/100b_t100tracer}
  \caption{\label{fig:100tracers} A visualization of the streamlines
  with 200 tracers passing through an eroded geometry with a porosity of
  0.63. The tracers are initialized at the left end of the channel and
  the flow is driven by a pressure difference from left to right.}
\end{wrapfigure}

@@

In an eroded geometry $\Omega$, the dimensionless
concentration $c$ is governed by
\begin{align}
  \pd{c}{t} + \uu \cdot \nabla c = \frac{1}{\Pe} \Delta c, 
  \label{eqn:advectionDiffusion}
\end{align}
where $\Pe$ is the dimensionless Peclet number. Here we assume that the
concentration $c$ is affected by the eroded geometry, but the geometry
is not altered further by the concentration. In the infinite Peclet
number limit, the concentration is governed by pure advection and the
characteristics of the concentration field are identical to the
streamlines (Fig.~\ref{fig:100tracers}).

The $j^{th}$ trajectory, or characteristic, is governed by $\dot{\ss}_j(t) = \uu(\ss_j(t))$. Once these trajectories are computed, using for example a high-order Runge-Kutta method, a statistical analysis can be used to describe the transport. Two of the most important measurements are the tortuosity and the anomalous dispersion, both which are defined in terms of the length of the trajectories \begin{align}
  \lambda_j(t) = \int_{0}^{t} \|\ss'(t')\| dt'.
\end{align}
Upon computing $N_p$ trajectories, the tortuosity is the dimensionless number
\begin{align}
  T = \frac{1}{d} \left(\int_{S} u_1(x_0,y_0)\lambda(y_0) dy_0\, \right)
  \Bigg/ \left(\int_S u_1(x_0,y_0)\, dy_0 \right),
\end{align}
where $S$ is the vertical cross-section at $x=x_0$, $d$ is its length,
$u_1$ is the horizontal velocity, and $\lambda(y_0)$ is the length of
the trajectory starting at $(x_0,y_0)$ and ending at a vertical
cross-section at the other end of the channel. Relationships between the
tortuosity and porosity of a geometry have been developed by other
groups~\cite{kop-kat-tim1996, dud-koz-mat2011, mat-kha-koz2008}, and
preliminary results indicate that such models might be appropriate for
geometries whose porosity is increasing under the process of erosion
(Fig.~\ref{fig:100tortuosity}(a)). However, preliminary results also
indicate that erosion can increase the amount of tortuosity as is
evident towards the at the towards the end of the erosion process in
Fig.~\ref{fig:100tortuosity}(a). Moreover, we have observed a
significant increase in the tortuosity at small porosities
(Fig.~\ref{fig:100tortuosity}(b)).

\begin{figure}[htp]
  \begin{center}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian100}
  \caption{100 eroding bodies}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian20}
  \caption{20 eroding bodies}
  \end{subfigure}
  \end{center}
  \caption{\label{fig:100tortuosity} The tortuosity of two different eroding geometries. The tortuosity can be computed using two different formulas (red vs.~blue marks). In the left plot, the red square is the tortuosity of the geometry in Fig.~\ref{fig:100tracers}. In both cases, there is qualitative agreement between our numerical simulations and a power-law model (black dashed line)~\cite{mat-kha-koz2008}.}
\end{figure}

While the tortuosity characterizes the amount of winding of streamlines, the dispersion characterizes the spreading of the trajectories.  Dispersion can be described using the first- and second-ensemble moments of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \lambda_j(t), \quad \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2
\end{align}
Then, the standard deviation $\sigma_\lambda$ characterizes the
dispersion, and long-time behavior typically results in a power-law
scaling $\sigma_\lambda \sim t^{\alpha}$. The power $\alpha$, which
quantifies the long-time spreading, is $1$ in a ballistic regime which
occurs at early times when the trajectories have not explored much of
the porous geometry. However, once the trajectories have traversed a few
grains, dispersion in porous media is often super-dispersive with
$\alpha \in (0.5,1)$~\cite{dea-leb-den-tar-bol-dav2013}. In Fig.~\ref{fig:100dispersion}, we investigate spreading at several different porosities of a porous geometry that is initialized with 100 grains.
\begin{wrapfigure}[22]{r}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} The particle spreading,
  characterized with the standard deviation $\sigma_\lambda(t)$ at 6
  different porosities. At early times, the line of slope 1 (black
  curve) indicates a ballistic motion. At later times, the
  dispersion becomes super-dispersive at a rate indicated by the slope
  of the black dashed-dotted line.}
\end{wrapfigure}

The rate of dispersion is known to be directly related to the pore size
distribution~\cite{dea-qua-bir-jua2018}. Therefore, there is a need to
understand how erosion affects the pore size distribution. In
preliminary work~\cite{chi-moo-qua2019}, we define pore sizes between
grains that share an edge of a Delanuay triangulation with vertices at
the grain centers (Fig.~\ref{fig:Eroding100gap_mean_var}(a)). In Fig.~\ref{fig:Eroding100gap_mean_var}(a) and~\ref{fig:Eroding100gap_mean_var}(b), we plot the mean and variance of the pore sizes as a function of porosity. To better understand how erosion affects the anomalous dispersion, we propose to investigate how the pores grow with erosion. We will start with the notion of a {\em porelet} which is an individual Hagen-Poiseuille flow in each of the individual pores~\cite{dea-qua-bir-jua2018}. By computing the strength of the individual Hagen-Poiseuille flows, the shear rate can be computed, and thus the rate of erosion can be computed.

We have performed a carefully chosen simulation to better understand the
effect of erosion on the pore size. We start with a single grain inside
the channel that is sufficiently large that the spacing between the
initial grain and the channel walls, which is 2 units wide), is
$10^{-3}$. Since we use a Barycentric quadrature rule, we are able to
stably simulate erosion with only 512 points on the eroding body which
corresponds to an arclength spacing 12 times larger than the original
separation distance---to achieve a comparable accuracy with the spectral
trapezoid rule, 60 times as many discretization points would be
required. In Fig.~\ref{fig:porelets}(a), different shades of grey
correspond to the configuration of the eroding grain at five equally
spaced time steps. At three of the time steps, we plot the velocity
between the grain and the solid wall (black line). At all three
configurations, the flow is nearly parabolic indicating that  porelet
model~\cite{dea-qua-bir-jua2018} is appropriate. After performing a
thorough analysis of the pore size growth in this simple configuration,
we will develop a more complete model for networks of pores that are
prevalent in porous media.

@@

\todo[inline]{Eventually, our FMM solver will reach its capacity. We
need to talk about fast summation methods for the barycentric
quadrature, deformation tensor, vorticity, and pressure. In addition to
fast summation methods, we can talk about parallel implementations. I
have new PhD student in DSC that wants to work on erosion and he has
experience running MPI and OpenMP.}

These simulations have not yet begun to incorporate the slow-acting effects of mechanical and chemical erosion, and the associated gradual changes in medium properties. A grand goal of this proposal would be to merge the two lines of inquiry to develop a sound theoretical framework for porous-medium dynamics robust enough to capture both short-term events, such as sinkhole collapse, and long-term changes in the medium. The greatest challenge faced is the disparate temporal and spatial scales that must be considered. Our proposed strategy is to use the high-fidelity erosion simulations to parametrize the effect of erosion on medium properties.

@@

\subsection{Undergraduate research}
Regarding undergraduate research, the ideas in this proposal represent a rich and fascinating topic for undergraduate students, while also being approachable these students. In particular, we aim to recruit at least one undergraduate student to perform laboratory experiments at the Geophysical Fluid Dynamics Institute (GFDI). As mentioned above, GFDI offers state-of-the-art laboratory facilities, with an already-functioning sinkhole experiment. The student will perform controlled experiments using natural and artificial porous media (e.g.~sand and micro-glass beads respectively) to precisely characterize the parameter regime leading to gravitational collapse. Parallel experiments will be conducted with erodible (bentonite clay) or dissolvable  (sucrose solids) fragments materials, which will represent the porous medium. These experiments will be used to guide, calibrate, and validate the computational advances. These undergraduate students will work closely with the graduate students to ensure that theory and experiments advance in close cooperation, with constant feedback between the two.

PI Moore already has experience successfully training undergraduate students in combined laboratory and theoretical work. A recent example is undergraduate student Tyler Bolles who completed an honors thesis with PI Moore at FSU on laboratory and theory of water waves. This work resulted in one {\em peer-reviewed} publication already~\cite{Bolles2019}, with a second to be submitted soon. In addition, PI Quaife successfully trained an undergraduate student, Miles Currie, in a statistical analysis of flows related to prescribed fire dynamics.  This work resulted in a peer-reviewed publication~\cite{cur-spe-hie-obr-goo-qua2018}.

@@

\subsection{Outreach for K--12 and the general public}
The PIs have already seized the opportunity to engage K--12 students in outreach activities. In particular, PI Moore's research was the focus of an educational video created by {\bf CPALMS}, which is the State of Florida's official source for standards information and course descriptions for K--12 education. The video, which features an interview and animations of research, is used to reinforce concepts from mathematics courses (grade levels 7--12), and to encourage students to consider a career in the STEM fields~\cite{CPALMS}. Additionally, these videos are frequently used for the continuing education of K--12 {\em teachers}, providing them a broader perspective of mathematics that could be integrated into their teaching.

@@

FSU also provides an opportunity to expose high school students to university-level research through the {\bf Young Scholars Program (YSP)}. The YSP is a six week experience that brings Florida's brightest grade 11 students to FSU where they are exposed to university-level lectures and work with FSU faculty on a research project. PI Quaife participated in this program in Summer 2016. The PIs will propose research projects that are accessible to high school students throughout the duration of the proposed research. In addition to high school students receiving a high school experience, the PhD students will be tasked with much of the day-to-day interaction with the high school students. This offers an excellent opportunity for the PhD students to develop leadership and mentorship skills.
